---
title: "Computation Musciology: Week 8"
author: "Dante van der Heijden"
date: "20 February 2019"
output: 
    flexdashboard::flex_dashboard:
        storyboard: true
        theme: lumen
---

```{r}
# Load required libraries, credentials and dataframes
if(!exists("utilize", mode="function")){ 
  source("./util.R")
  utilize()
}

```

### The Neptunes Sound in Popular Music

To understand what role The Neptunes played in the rise of Hip-Hop and R&B-influenced music in the pop charts, a corpus consisting of 798 songs made by The Neptunes and later solo by Pharrell Williams (playlist by Dante van der Heijden) is analyzed and compared to other popular music in the 2000’s to see see where the music by The Neptunes fitted in popular music during the mid 2000’s, and how it may have shaped what is played on the radio since then. To address this comparison, the “2000 pop Hits” playlist by QuLa Khan is analyzed, containing 231 pop hits from 2000 - 2010.

## Background

Pharrell Williams is most notably known for his recent works on the minions soundtrack (“Happy”) and the work he made with his band, N.E.R.D. Pharrell’s biggest success however, was during the mid 2000’s when he and his counterpart Chad Hugo produced hundreds of songs as “The Neptunes”. Starting off in 1992 with the US #2 Single “Rump Shaker” by Wreckx-N-Effect; the typical Neptunes style was first heard: rhythmical beats and a lot of danceability. This is also reflected by the Spotify data, giving the 1992 song a danceability index of 0.865. The typical four count start, which ended up being Pharrell’s musical signature, was first heard on the 1996 song “When Boy Meets Girl” by Total. The introduction of what would be regarded as “The Neptunes Sound” came when “Superthug” was released by “N.O.R.E.”: synthesizers, keyboard sampling and modules. Even more commercial success followed, with their first worldwide hit in 2001 with Britney Spears’ “I’m a Slave 4 U”. Pharrell Williams has always done some solo production work, but starting in 2010 with the original Despicable Me soundtrack Pharrell started to focus more on his solo career, with productions for artists like Frank Ocean, Kendrick Lamar, Tyga, Pusha T and Beyoncé. 

## Preliminary Findings

When solely looking at the data, one figure immediately stands out: danceability. It’s mean is 0.735 with a standard error of 0.130. The energy is also really present, with a mean of 0.704 and a standard error of 0.152. It also shows that the songs produced by Pharrell and Chad are regarded as positive, with a mean valance of 0.651 (s.e. of 0.204). These three attributes for the Pop 2000 playlist show lower figures (s.e.’s in brackets): 0.652 (0.145), 0.712 (0.152), 0.572 (0.219). The energy attribute is the only attribute that scores higher in the Pop 2000 playlist. 

When looking at the extreme values of the Neptunes playlist, Spotify seemed to have done a good job on indexing the danceability, energy and valance. The same holds for the Pop playlist. 


### Some nice graphs

```{r}
# Filtering the data
## For Pharrell
pharrell_danceability_by_year <- pharrell %>%
  mutate(year = format(release_date, "%Y")) %>%
  group_by(year) %>% 
  filter(year >= 1992) %>%
  summarise(dance = mean(danceability)*100, pop = mean(track_popularity), songs = n())

pharrell_danceability_popularity <- pharrell %>%
  mutate(track_popularity = track_popularity/max(track_popularity)*100) %>% 
  mutate(year = format(release_date, "%Y")) %>%
  group_by(year) %>% filter(year >= 1992)

## For Pop2000
pop2000_danceability_by_year <- pop2000 %>%
  mutate(year = format(release_date, "%Y")) %>%
  group_by(year) %>%
  summarise(dance = mean(danceability)*100, pop = mean(track_popularity), songs = n())

pop2000_danceability_popularity <- pop2000 %>%
  mutate(track_popularity = track_popularity/max(track_popularity)*100) %>% 
  mutate(year = format(release_date, "%Y")) %>%
  group_by(year) %>% filter(year >= 1992)

# Plotting the graphs
## For Pharrell
ggplot(pharrell_danceability_by_year, aes(year, dance, group=1)) +
  geom_point(aes(size = songs, color = pop)) +
  scale_color_distiller(palette = "Spectral") +
  xlab("Year") +
  ylab("Dancebility") +
  labs(size="Amount of Songs", colour = "Popularity") +
  scale_x_discrete(breaks = scales::pretty_breaks(n = 5))

ggplot(pharrell_danceability_popularity, aes(danceability, track_popularity)) +
  geom_point() +
  geom_smooth(method=lm)

## For Pop2000
ggplot(pop2000_danceability_by_year, aes(year, dance, group=1)) +
  geom_point(aes(size = songs, color = pop)) +
  scale_color_distiller(palette = "Spectral") +
  xlab("Year") +
  ylab("Dancebility") +
  labs(size="Amount of Songs", colour = "Popularity") +
  scale_x_discrete(breaks = scales::pretty_breaks(n = 5))

ggplot(pop2000_danceability_popularity, aes(danceability, track_popularity)) +
  geom_point() +
  geom_smooth(method=lm)
```

***

Here is a thorough description of the visualisation.
